Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-09-09 submission/manuscript.Rmd
render figure legends and fix plus minus issue

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/KDM5C_ChIPseq.pdf}
  \caption[KDM5C binds to a subset of germline gene promoters during early embryogenesis]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis A.} ChIPseeker localization of KDM5C peaks at different genomic regions in EpiLCs (top) and hippocampal and cortex primary neuron cultures (PNCs, bottom). \textbf{B.} Overlap of genes with KDM5C bound to their promoters (TSS &#177; 500) in EpiLCs (purple) and PNCs (blue). \textbf{C.} Gene ontology (GO) comparision of genes with KDM5C bound to their promoter in EpiLCs and PNCs. Genes were classified as either bound in EpiLCs only (EpiLC only), unique to PNCs (PNC only, no significant ontologies) or bound in both PNCs and EpiLCs (shared). \textbf{D.} Average KDM5C binding around the transcription start site (TSS) of all germline-enriched genes in EpiLCs (left) and PNCs (right). \textbf{E.} Eulerr of number of germline-enriched genes (green) with significant KDM5C binding at their promoter in EpiLCs (purple). \textbf{F.} Example KDM5C ChIP-seq bigwigs of KDM5C binding at the \textit{Dazl} TSS but not \textit{Stra8} in EpiLCs. {G.} HOMER motif analysis of all KDM5C-bound germline gene promoters, highlighting significant enrichment of MAX, E2F6, and E2F motifs. \textbf{H.} Number of all gene promoters bound or unbound by KDM5C with instances of the E2F or E-box consensus sequence. \textbf{I.} HOMER motif analysis of all KDM5C-unbound germline gene promoters, highlighting significant enrichment of RFX family transcription factor motifs. \textbf{J.} Expression of RNA-seq DEG \textit{Rfx2} in wild-type and \textit{Kdm5c}-KO EpiLCs. P-value of Welch's t-test, expression in transcripts per million (TPM).}
  \label{figurelabel}
\end{figure}

\begin{figure}
  \centering
  \includegraphics{../submission/compiled_figs/KDM5C_Mechanism.pdf}
  \caption[KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation]{\textbf{KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation.} \textbf{A.} Left - Bigwigs of representative histone 3 lysine 4 trimethylation (H3K4me3) ChIP-seq peaks at two germline genes in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) adult amygdala. Right - Average H3K4me3 at the transcripiton start site (TSS) of all germline-enriched genes in wild-type (dark red) and \textit{Kdm5c}-KO (light red) amygdala. \textbf{B.} Left - Bigwigs of representative histone 3 lysine 4 dimethylation (H3K4me2) ChIP-seq peaks at representative germline genes in wild-type and \textit{Kdm5c}-KO EpiLCs. Right - Average H3K4me2 at the TSS of all germline-enriched genes in wild-type (dark purple) and \textit{Kdm5c}-KO (light purple) EpiLCs. \textbf{C.} Diagram of embryonic stem cell (ESC) to epiblast-like cell (EpiLC) differentiation protocol and collection time points for RNA and protein. \textbf{D.} Real time quantitative PCR (RT-qPCR) of \textit{Kdm5c} mRNA expression, calculated in comparision to TBP expression (2\textsuperscript{-deltaCT}). * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test. \textbf{E.} KDM5C protein expression normalized to DAXX. Quantified intensity using ImageJ (artificial units - au). Right - representative lanes of Western blot for KDM5C and DAXX. * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test \textbf{F.} Percentage of germline genes that harbor CpG islands (CGIs) in their promoters, based on UCSC annotation. Comparing all germline-enriched genes, KDM5C-bound germline genes, or KDM5C-unbound germline genes \textbf{G.} enrichPlot gene ontology analysis of CGI-promoter versus non-CGI promoter germline genes. \textbf{H.} Diagram of ESC to extended EpiLC (exEpiLC) differentiation \textbf{I.} Volcano plot of whole genome bisulfite sequencing (WGBS) comapring CpG methylation at germline gene promoters (TSS &#177; 500) in wild-type ESCs versus exEpiLCs. Promoter CGI genes highlighted in red, hypermethylated genes lackig a promoter CGI in pink \textbf{J.} Volcano plot of WGBS of wild-type versus \textit{Kdm5c}-KO exEpiLCs. Promoter CGI genes highlighted in red, hypermethylated genes lackig a promoter CGI in pink, hypomethylated genes lackig a promoter CGI in blue. \textbf{K.} Example UCSC browser shots of germline genes of CpG methylation (CpGme) in wild-type and \textit{Kdm5c}-KO ESCs and exEpiLCs. \textbf{L.} Histogram of average percent CpGme at the promoter for germline genes with or without promoter CGIs. Wild-type in navy and \textit{Kdm5c}-KO in light blue. Dashed lines are average methylation for each genotype, p-values for Mann−Whitney U test.}
  \label{figurelabel}
\end{figure}

>>>>

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/KDM5C_ChIPseq.pdf}
  \caption[KDM5C binds to a subset of germline gene promoters during early embryogenesis.]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis.} \textbf{A.}  ChIPseeker localization of KDM5C peaks at different genomic regions in EpiLCs (top) and hippocampal and cortex primary neuron cultures (PNCs, bottom). \textbf{B.} Overlap of genes with KDM5C bound to their promoters (TSS $\pm$ 500) in EpiLCs (purple) and PNCs (blue). \textbf{C.} Gene ontology (GO) comparision of genes with KDM5C bound to their promoter in EpiLCs and PNCs. Genes were classified as either bound in EpiLCs only (EpiLC only), unique to PNCs (PNC only, no significant ontologies) or bound in both PNCs and EpiLCs (shared). \textbf{D.} Average KDM5C binding around the transcription start site (TSS) of all germline-enriched genes in EpiLCs (left) and PNCs (right). \textbf{E.} Eulerr of number of germline-enriched genes (green) with significant KDM5C binding at their promoter in EpiLCs (purple). \textbf{F.} Example KDM5C ChIP-seq bigwigs of KDM5C binding at the \textit{Dazl} TSS but not \textit{Stra8} in EpiLCs. {G.} HOMER motif analysis of all KDM5C-bound germline gene promoters, highlighting significant enrichment of MAX, E2F6, and E2F motifs. \textbf{H.} Number of all gene promoters bound or unbound by KDM5C with instances of the E2F or E-box consensus sequence. \textbf{I.} HOMER motif analysis of all KDM5C-unbound germline gene promoters, highlighting significant enrichment of RFX family transcription factor motifs. \textbf{J.} Expression of RNA-seq DEG \textit{Rfx2} in wild-type and \textit{Kdm5c}-KO EpiLCs. P-value of Welch's t-test, expression in transcripts per million (TPM).}
  \label{figurelabel}
\end{figure}

\begin{figure}
  \centering
  \includegraphics{../submission/compiled_figs/KDM5C_Mechanism.pdf}
  \caption[KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation]{\textbf{KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation.} \textbf{A.} Left - Bigwigs of representative histone 3 lysine 4 trimethylation (H3K4me3) ChIP-seq peaks at two germline genes in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) adult amygdala. Right - Average H3K4me3 at the transcripiton start site (TSS) of all germline-enriched genes in wild-type (dark red) and \textit{Kdm5c}-KO (light red) amygdala. \textbf{B.} Left - Bigwigs of representative histone 3 lysine 4 dimethylation (H3K4me2) ChIP-seq peaks at representative germline genes in wild-type and \textit{Kdm5c}-KO EpiLCs. Right - Average H3K4me2 at the TSS of all germline-enriched genes in wild-type (dark purple) and \textit{Kdm5c}-KO (light purple) EpiLCs. \textbf{C.} Diagram of embryonic stem cell (ESC) to epiblast-like cell (EpiLC) differentiation protocol and collection time points for RNA and protein. \textbf{D.} Real time quantitative PCR (RT-qPCR) of \textit{Kdm5c} mRNA expression, calculated in comparision to TBP expression (2\textsuperscript{-deltaCT}). * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test. \textbf{E.} KDM5C protein expression normalized to DAXX. Quantified intensity using ImageJ (artificial units - au). Right - representative lanes of Western blot for KDM5C and DAXX. * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test \textbf{F.} Percentage of germline genes that harbor CpG islands (CGIs) in their promoters, based on UCSC annotation. Comparing all germline-enriched genes, KDM5C-bound germline genes, or KDM5C-unbound germline genes \textbf{G.} enrichPlot gene ontology analysis of CGI-promoter versus non-CGI promoter germline genes. \textbf{H.} Diagram of ESC to extended EpiLC (exEpiLC) differentiation \textbf{I.} Volcano plot of whole genome bisulfite sequencing (WGBS) comapring CpG methylation at germline gene promoters (TSS $\pm$ 500) in wild-type ESCs versus exEpiLCs. Promoter CGI genes highlighted in red, hypermethylated genes lackig a promoter CGI in pink \textbf{J.} Volcano plot of WGBS of wild-type versus \textit{Kdm5c}-KO exEpiLCs. Promoter CGI genes highlighted in red, hypermethylated genes lackig a promoter CGI in pink, hypomethylated genes lackig a promoter CGI in blue. \textbf{K.} Example UCSC browser shots of germline genes of CpG methylation (CpGme) in wild-type and \textit{Kdm5c}-KO ESCs and exEpiLCs. \textbf{L.} Histogram of average percent CpGme at the promoter for germline genes with or without promoter CGIs. Wild-type in navy and \textit{Kdm5c}-KO in light blue. Dashed lines are average methylation for each genotype, p-values for Mann-Whitney U test.}
  \label{figurelabel}
\end{figure}
